---
title: "Simulation and Analysis Project"
author: "Nicole Scott"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 8)
```


##Simulation

###Overview

This simulation investigates the distribution of the mean of 40 numbers drawn
from an exponential distribution with rate = 0.2.  Though the simulated data
does not follow a normal distribution, the distribution of means can be 
approximated by the normal distribution with mean = 1/rate and standard of
deviation = 1/rate.  This follows the Central Limit Theorem.

###Simulation

```{r simulation}
lambda <- 0.2
sample <- rexp(40*1000, rate = lambda)
samplemat <- matrix(sample, nrow = 1000, ncol = 40)

```

This generates data points from an exponential distribution with rate lambda = 
`r lambda`.  Data was generated for 1000 sets of 40 points and then arranged in 
a matrix with each row corresponding to one sample set.

```{r simulation plot}
library(ggplot2)
ggplot() +
        geom_histogram(aes(sample, y = ..density..), color = "grey",
                       fill = "steelblue", bins = 20)+
        geom_density(aes(sample)) +
        xlab("") +
        ggtitle("Simulated Exponential Data with Rate = 0.2")
```

This plot is a histogram of the simulated data, overlayed with a density curve.

###Sample Mean versus Theoretical Mean

```{r means}
samplemeans <- rowMeans(samplemat)
samplemean <- round(mean(samplemeans), 3)
theorymean <- 1/lambda
ggplot()+
        geom_histogram(aes(samplemeans, y = ..density..), color = "grey", 
                       fill = "steelblue", bins = 20) +
        geom_vline(xintercept = samplemean) +
        xlab("Mean of 40 exponentials with lambda = 0.2")+
        ggtitle("Distribution of the Means of 40 Exponentials") +
        geom_density(aes(samplemeans))

```

First, row means were calculated for the generated matrix to generate means of
40 samples each and stored as samplemeans.  The plot is a histogram of these 
means, with a vertical line indicating the average of the mean, rounded to 3 
decimal places(`r samplemean`), with an overlayed density curve. Theoretically 
the mean of exponential data would be 1/rate, in this case `r theorymean`.  The 
difference between the sample mean (for means of 40 over 1000 simulations) and 
the theoretical mean is `r samplemean - theorymean`.

###Sample Variance versus Theoretical Variance

```{r variance}
samplesd <- round(sd(sample), 3)
meansd <- round(sd(samplemeans), 3)
theorysd <- 1/lambda
```

The standard deviation (sigma) was calculated for the means to approximate the 
variability in means of 40 exponentials. The theoretical standard deviation 
is also equal to 1/rate.  The difference between the theoretical sigma and the 
sample sigma is `r theorysd - samplesd`.  However, the difference between 
the theoretical sigma of an exponential distribution and the sigma of means of 
40 exponentials is `r theorysd - meansd`.  (The difference between the 
variances (sigma^2) is `r theorysd^2 - meansd^2`).  The variance of the means
is smaller than the variance of the population, following the Central Limit 
Theorem, as the distribution narrows around the population mean.


###Distribution

```{r distribution plot}
ggplot(data.frame(sample), aes(sample)) +
        geom_histogram(aes(sample, y = ..density..), color = "grey", fill = "steelblue", position = "stack", bins = 20)+
        geom_density(aes(sample, col = "Sample Distribution")) +
        stat_function(fun = dnorm, 
                      args = list(mean = samplemean, sd = samplesd), 
                      geom = "line", aes(col = "Normal Distribution")) + 
        xlab("") +
        scale_color_manual(values = c("red", "black"), 
                           name = "") +
        ggtitle("Simulated Exponential Data with Rate = 0.2")
```

This plot shows the density of the simulated data compared to a normal curve
with mean = `r mean(sample)` and standard of deviation = `r sd(sample)`. The 
normal distribution does not approximate the sample population as a whole very 
well, which is logical as the samples were drawn from an exponential 
distribution, not a normal one. *As mean is a linear funciton, the mean was not
recalculated for the sample as a whole in comparison to the mean of the set of 
means.*

```{r dist means plot}
ggplot(data.frame(samplemeans), aes(samplemeans)) +
        geom_histogram(aes(samplemeans, y = ..density..), color = "grey", 
                       fill = "steelblue", position = "stack", bins = 20) +
        geom_density(aes(samplemeans, color = "Sample Mean Distribution")) +
        stat_function(fun = dnorm, 
                      args = list(mean = samplemean, sd = meansd), 
                      geom = "line", aes(col = "Normal Distribution")) + 
        scale_color_manual(values = c("red", "black"),
                           name = "") +
        xlab("Mean") +
        ggtitle("Means of 40 Exponentials with Rate = 0.2")
```

The distribution of the means of sets of 40 sampled points is much better 
approximated by the normal curve, following the central limit theorem.  

===============================================================================

##Tooth Growth Data Analysis

###Exploratory Analysis and Data Summary

```{r exploratory}
library(datasets)
str(ToothGrowth)
summary(ToothGrowth)
ggplot(ToothGrowth) +
        geom_boxplot(aes(x = supp, y = len)) +
        ylab("Tooth Length") +
        xlab("Vitamin C Supplement Type (OJ = Orange Juice, 
             VC = Absorbic Acid)") +
        ggtitle("The Effect of Vitamin C on Tooth Growth in Guinea Pigs")

```

This data set reports the length of odontoblasts (cells responsible for tooth 
growth) in 60 guinea pigs after different doses of Vitamin C given through
orange juice or absorbic acid supplements (as described in the documentation).
Using the str() function, it was found that the dataframe contained 3 variables
for each of the 60 guinea pigs, "len" - a numeric variable for tooth length -
"supp" - a factor indicating the type of supplement - and "dose" - a numeric
variable corresponding to one of the 3 doses (0.5, 1, or 2 mg/day).  Summary()
was used to get an overview of the values in the dataset.  The boxplot shows
the difference between the two supplement types regardless of dose, with orange
juice appearing to result in higher tooth length values.

```{r exploratory2}
ggplot(ToothGrowth) +
        geom_boxplot(aes(x = supp, y = len, fill = factor(dose)))+
        ylab("Tooth Length") +
        xlab("Vitamin C Supplement Type (OJ = Orange Juice, VC = Absorbic Acid)") +
        labs(fill = "Dose (mg/day)") +
        ggtitle("The Effect of Vitamin C on Tooth Growth in Guinea Pigs")

```

This plot further breaks down the previous boxplot by dose.  It appears within
each supplement type, tooth length increases with increased doseage.  It also
appears that orange juice also resulted in higher tooth length values within 
each dose, though that is less clear when the dose is 2 mg/day.  These are the
trends that will be investigated in the analysis of the data.  

###Analysis

*For all comparisons in this analysis the test statistic used will be the mean.*
*None of the comparisons will be considered paired as each observation* 
*corresponds to a different guinea pig.*

*Supplement Type Comparison*

First to be examined is the hypothesis that orange juice supplements result in
higher tooth length values overall (using values from all doses).  

muo = average of OJ supplement group
muv = average of absorbic acid group

H0: muo = muv
Ha: muo > muv

```{r supp hypothesis}

```

The next hypotheses examined is that increased dose groups result in a higher 
mean tooth length.  Specifically the confidence interval for the middle dose 
for each supplement type will be compared to the higher and lower doses to 
determine if the patterns shown in the exploratory graphs are statistically 
significant.

*Orange Juice Group by Dose*

muo.5 = mean orange juice 0.5 mg dose group
muo1 = mean orange juice 1 mg dose group
muo2 = mean orange juice 2 mg dose group

*Absorbic Acid Group by Dose*

muo.5 = mean absorbic acid 0.5 mg dose group
muo1 = mean absorbic acid 1 mg dose group
muo2 = mean absorbic acid 2 mg dose group

###Conclusions