---
title: "Simulation and Analysis Project"
author: "Nicole Scott"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 8)
```


##Simulation

###Overview

This simulation investigates the distribution of the mean of 40 numbers drawn
from an exponential distribution with rate = 0.2.  Though the simulated data
does not follow a normal distribution, the distribution of means can be 
approximated by the normal distribution with mean = 1/rate and standard of
deviation = 1/rate.  This follows the Central Limit Theorem.

###Simulation

```{r simulation}
lambda <- 0.2
sample <- rexp(40*1000, rate = lambda)
samplemat <- matrix(sample, nrow = 1000, ncol = 40)

```

This generates data points from an exponential distribution with rate lambda = 
`r lambda`.  Data was generated for 1000 sets of 40 points and then arranged in 
a matrix with each row corresponding to one sample set.

```{r simulation plot}
library(ggplot2)
ggplot() +
        geom_histogram(aes(sample, y = ..density..), color = "grey",
                       fill = "steelblue", bins = 20)+
        geom_density(aes(sample)) +
        xlab("") +
        ggtitle("Simulated Exponential Data with Rate = 0.2")
```

This plot is a histogram of the simulated data, overlayed with a density curve.

###Sample Mean versus Theoretical Mean

```{r means}
samplemeans <- rowMeans(samplemat)
samplemean <- round(mean(samplemeans), 3)
theorymean <- 1/lambda
ggplot()+
        geom_histogram(aes(samplemeans, y = ..density..), color = "grey", 
                       fill = "steelblue", bins = 20) +
        geom_vline(xintercept = samplemean) +
        xlab("Mean of 40 exponentials with lambda = 0.2")+
        ggtitle("Distribution of the Means of 40 Exponentials") +
        geom_density(aes(samplemeans))

```

First, row means were calculated for the generated matrix to generate means of
40 samples each and stored as samplemeans.  The plot is a histogram of these 
means, with a vertical line indicating the average of the mean, rounded to 3 
decimal places(`r samplemean`), with an overlayed density curve. Theoretically 
the mean of exponential data would be 1/rate, in this case `r theorymean`.  The 
difference between the sample mean (for means of 40 over 1000 simulations) and 
the theoretical mean is `r samplemean - theorymean`.

###Sample Variance versus Theoretical Variance

```{r variance}
samplesd <- apply(samplemat, 1, sd)
meansd <- round(mean(samplesd), 3)
theorysd <- 1/lambda
ggplot()+
        geom_histogram(aes(samplesd, y = ..density..), color = "grey", 
                       fill = "steelblue", bins = 20)+
        geom_vline(xintercept = meansd) + 
        geom_density(aes(samplesd))+
        xlab("Standard of deviation of 40 exponentials with lambda = 0.2")+
        ggtitle("Distribution of the Variance of 40 Exponentials")
```

First, the standard deviation (sigma) was calculated row-wise for each of the 
sets of 40 exponentials.  The plot is a histogram of these standard 
deviations with a vertical line at the mean of the test stat rounded to 3 
decimal places, `r meansd`.  The theoretical standard deviation is also equal 
to 1/rate.  The difference between the sample sigma and the theoretical value is 
`r meansd - theorysd`.  The difference of the variances (sigma^2) is equal to
`r meansd^2 - theorysd^2`.

###Distribution

```{r distribution plot}
ggplot(data.frame(sample), aes(sample)) +
        geom_histogram(aes(sample, y = ..density..), color = "grey", fill = "steelblue", position = "stack", bins = 20)+
        geom_density(aes(sample, col = "Sample Distribution")) +
        stat_function(fun = dnorm, 
                      args = list(mean = samplemean, sd = sd(sample)), 
                      geom = "line", aes(col = "Normal Distribution")) + 
        xlab("") +
        scale_color_manual(values = c("red", "black"), 
                           name = "") +
        ggtitle("Simulated Exponential Data with Rate = 0.2")
```

This plot shows the density of the simulated data compared to a normal curve
with mean = `r samplemean` and standard of deviation = `r sd(sample)`. Because
mean is a linear function, it was not recalculated for the total dataset (the
mean of the row-wise means = the mean of all data).  The normal distribution
does not approximate the sample population as a whole very well, which is 
logical as the samples were drawn from an exponential distribution.

```{r dist means plot}
ggplot(data.frame(samplemeans), aes(samplemeans)) +
        geom_histogram(aes(samplemeans, y = ..density..), color = "grey", fill = "steelblue", position = "stack", bins = 20)+
        geom_density(aes(samplemeans, color = "Sample Mean Distribution")) +
        stat_function(fun = dnorm, 
                      args = list(mean = samplemean, sd = sd(samplemeans)), 
                      geom = "line", aes(col = "Normal Distribution")) + 
        scale_color_manual(values = c("red", "black"),
                           name = "") +
        xlab("Mean") +
        ggtitle("Means of 40 Exponentials with Rate = 0.2")
```

The distribution of the means of sets of 40 sampled points is much better 
approximated by the normal curve, following the central limit theorem.  

===============================================================================

##Tooth Growth Data Analysis

###Exploratory Analysis

###Data Summary

###Analysis

###Conclusions